---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
---

<!-- This section is mandatory! -->
\captionsetup{width=6.5in} <!-- makes table caption stays in margins -->

### Differential expression

(ref:pilotPCA-cap) \textbf{Correlation of principal component analysis (PCA) components (PCs) with demographic and clinical metadata.} Values shown are Pearson correlation coefficients. 

```{r pilotPCA, eval = T, out.width='0.8\\linewidth',echo=FALSE, fig.cap="(ref:pilotPCA-cap)", fig.scap = "Correlation of PCs with metadata (pilot study)", out.extra='', fig.align='center'}
knitr::include_graphics("./Figures/BloodPilot/Figure E1 PCA metadata.pdf")
```

\newpage
\begin{singlespace}

(ref:pilotpftdeg-cap) \textbf{List of transcripts analyzed for differential expression with predicted forced vital capacity (FVC\%) and diffusing capacity of the lungs for carbon monoxide (D\textsubscript{LCO}\%).} Results shown are adjusted for treatment status and race. 

```{r pilotpftdeg, eval = T, echo = F, warning = FALSE, message=FALSE, out.extra='', fig.scap='ILD pilot PFT DEGs', fig.align='center'}
deg_pft_table <- readr::read_csv(here::here('Tables', 'BloodPilot', 'diffexppft.csv'))

# function for sig figs
round_to_sigfigs <- function(x, sigfigs) {
  if (!is.numeric(x)) return(x)  
  round(x, digits = sigfigs)
}
deg_pft_table <- deg_pft_table %>%
  mutate_if(is.numeric, round_to_sigfigs, sigfigs = 3)

colnames(deg_pft_table) <- rep(c("Transcript","(ref:logfc)", "(ref:p-it)-value", "FDR"), 2)

kableExtra::kbl(
  deg_pft_table,
  longtable = T,
  format = 'latex',
  booktabs = T, 
  linesep = "",
  align = "lrcclrcc",
  caption="(ref:pilotpftdeg-cap)",
  caption.short = "Blood pilot study DEGs associated with PFT") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%  # table spans across left to right margins
  column_spec(c(1,5), width="0.8in") %>%
  column_spec(c(2,6), width="0.6in") %>%
  column_spec(c(3,7), width="0.6in") %>%
  column_spec(c(4,8), width="0.6in") %>%
  add_header_above(header = c("FVC%" = 4, "DLCO%" = 4)) %>%
  #kableExtra::kable_styling(latex_options = "scale_down") #%>% 
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 9) # ensures the table isn't "floated" and pins it to the current location
```

\end{singlespace}
\pagebreak

### Biomarker panels

\begin{singlespace}

\captionsetup{width=6.5in} 

(ref:biomarkermodel-cap) \textbf{Classification model performance on cross-validation for selected learning algorithms.} Data are reported as mean accuracy ± standard error. 

```{r biomarkermodel, eval = T, echo = F, warning = FALSE, message=FALSE, out.extra='', fig.scap='ILD pilot biomarker panel', fig.align='center'}
biomarker_spec <- readr::read_csv(here::here('Tables', 'BloodPilot', 'Table biomarker model.csv'))
biomarker_spec <- biomarker_spec %>% 
  mutate(sPLSDA = gsub("\\+","±",sPLSDA)) %>% 
  mutate(`Elastic Net`=gsub("\\+","±",`Elastic Net`)) %>%
  mutate(`Random Forest`=gsub("\\+","±",`Random Forest`))

kableExtra::kbl(
  biomarker_spec,
  longtable = F,
  format = 'latex',
  booktabs = T, 
  linesep = "",
  align = "lccc",
  caption="(ref:biomarkermodel-cap)",
  caption.short = "Blood pilot study ILD classification model performance") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%  # table spans across left to right margins
  column_spec(1, width="1.3in") %>%
  column_spec(2, width="0.7in") %>%
  column_spec(3, width="0.7in") %>%
  column_spec(4, width="0.7in") %>%
  #row_spec(37:38, bold=T) %>%  
  kableExtra::kable_styling(full_width = F) %>%
  #kableExtra::kable_styling(latex_options = "scale_down") #%>% 
  kableExtra::kable_styling(latex_options = c("HOLD_position", "repeat_header"), font_size = 8)  # ensures the table isn't "floated" and pins it to the current location
```

(ref:biomarkerweight-cap) \textbf{Coefficient weights of top 20 features (transcripts) of elastic net biomarker panels developed for each between-subtype comparison.} 

```{r biomarkerweight, eval = T, echo = F, warning = FALSE, message=FALSE, out.extra='', fig.scap='ILD pilot biomarker panel weights', fig.align='center'}
biomarker_weight <- readr::read_csv(here::here('Tables', 'BloodPilot', 'Table biomarker panel weights.csv'))

biomarker_weight <- as.data.frame(lapply(biomarker_weight, function(x) {
  gsub("_isoform", "", x)
}))
biomarker_weight <- as.data.frame(lapply(biomarker_weight, function(x) {
  gsub("_intron1.1", "", x)
}))

biomarker_weight <- biomarker_weight %>% mutate(across(c(2,4,6,8), as.numeric))

# function for sig figs
round_to_sigfigs <- function(x, sigfigs) {
  if (!is.numeric(x)) return(x)  
  round(x, digits = sigfigs)
}

biomarker_weight <- biomarker_weight %>%
  mutate_if(is.numeric, round_to_sigfigs, sigfigs = 3)

colnames(biomarker_weight) <- rep(c("Transcript","Weight"), 4)

kableExtra::kbl(
  biomarker_weight,
  longtable = T,
  format = 'latex',
  booktabs = T, 
  #linesep = "",
  align = "lrlrlrlr",
  caption="(ref:biomarkerweight-cap)",
  caption.short = "Blood pilot study ILD classification model weights") %>%
  add_header_above(header = c("IPF vs non-IPF" = 2, "IPF vs SSc-ILD" = 2, "IPF vs HP" = 2, "HP vs SSc-ILD" = 2)) %>%
  kableExtra::kable_styling(full_width = TRUE) %>%  # table spans across left to right margins
  column_spec(c(1,3,5,7), width="0.7in") %>%
  column_spec(c(2,4,6,8), width="0.6in") %>%
  #row_spec(37:38, bold=T) %>%
  kableExtra::kable_styling(full_width = F) %>%
  #kableExtra::kable_styling(latex_options = "scale_down") #%>% 
  kableExtra::kable_styling(latex_options = c("HOLD_position", "repeat_header"), font_size = 8)  # ensures the table isn't "floated" and pins it to the current location
```

\end{singlespace}
