---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
---

<!-- This section is mandatory! -->

The results from the previous two chapters highlight the link between transcriptomic profiles, mechanisms of ILD, and clinical variables. Given the heterogeneity within ILD subtypes, analysis of the peripheral transcriptome should consider radiological pattern, progressive phenotype, and other related clinical factors [@ryerson_lumpers_2019]. In this chapter, we will explore the peripheral transcriptome using RNA-seq data from whole blood samples of patients with ILD. 

## Methods

### Study population and sample processing

As part of the Canadian Registry for Pulmonary Fibrosis (CARE-PF), patients with ILD were prospectively recruited with written informed consent between 2012 and 2019 at the Pacific Lung Health Centre located at St. Paul's Hospital (Vancouver, BC, Canada). Demographic (age, race, sex) and clinical data (FVC%, D\textsubscript{LCO}% predicted) were collected at time of sampling (Table \@ref(tab:bloodpxrnaseq). ILD diagnoses were determined by consensus agreement by a MDD panel consisting of at least one pulmonologist, radiologist, and pathologist, and independently re-evaluated to confirm diagnoses and other clinical features (including HRCT patterns) between January 2021 and March 2022 [@marinescu_integration_2023]. Patients were defined as having PPF on the basis of the INBUILD criteria (see: [Introduction - Progressive pulmonary fibrosis](#ppf)). 

At time of recruitment, patient whole blood samples were collected in PAXgene Blood RNA Tubes (BD Biosciences, Mississauga, ON, Canada). We selected samples to examine differences between patients with ILD that experience PPF (n=39) and those that did not (n=57) with a balanced representation of ILD subtypes, and included samples if patients were not undergoing active pharmalogical treatment at time of blood sampling. PAXgene-preserved whole blood samples were extracted for RNA using the PAXgene Blood miRNA Kit (PreAnalytiX, Hombrechtikon, Switzerland), and quality control was performed using the Agilent 2100 Bioanalyzer (Agilent, Santa Clara, California, USA). Sample library preparation was performed using the Illumina Stranded mRNA Prep kit (Illumina, San Diego, California, USA), then sequenced on the Illumina NextSeq 2000 sequencing system with paired end 59bp x 59bp reads. Due to hemoglobin-associated genes comprising a majority of reads in whole blood transcriptome sequencing [@shin_variation_2014], each sample was sequenced to a depth a 50 million reads. Sequencing data quality check was performed using fastqc (v0.11.9), and sequence alignment to the human reference genome (GENCODE v41) was performed using STAR (v2.7.10a). Transcript abundance was estimated using featureCounts (v2.0.3). Hemoglobin genes (\textit{HBA1}, \textit{HBA2}, \textit{HBB}) were removed from the final count matrix due to their overrepresentation. 

\begin{landscape}

\captionsetup{width=8.5in}

(ref:race) \textbf{Race, n (\%)}

(ref:bloodpxrnaseq-cap) \textbf{Summary table of demographics of patients with ILD examined for whole blood RNA expression.} Data are shown as either n(\%) or mean ± SD (calculated for all patients within diagnoses unless otherwise indicated). \textit{P}-values were determined using one-way analysis of variance (ANOVA) for continuous variables and Fisher's exact text for categorical variables. Variables relating to progression were determined within a 2-year period. 

```{r bloodpxrnaseq, eval = T, echo = F, warning = FALSE, message=FALSE, out.extra='', fig.scap='ILD blood RNA-seq study cohort', fig.align='center'}
bloodpxrnaseq <- readr::read_csv(here::here('Tables', 'BloodRNAseq', 'px_summary.csv'))
colnames(bloodpxrnaseq)[1] <- ""
kableExtra::kbl(
  bloodpxrnaseq,
  longtable = F,
  format = 'latex',
  booktabs = T, 
  linesep = "",
  align = "lccccccl",
  caption="(ref:bloodpxrnaseq-cap)") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%  # table spans across left to right margins
  column_spec(1, width="1.5in") %>%
  column_spec(2, width="1.05in") %>% #SSc
  column_spec(3, width="1.05in") %>% #IPF
  column_spec(4, width="0.97in") %>% #HP
  column_spec(5, width="0.97in") %>% #RA
  column_spec(6, width="0.95in") %>% # IPAF
  column_spec(7, width="1.3in") %>% # Unc
  column_spec(8, width="0.7in") %>% # p
  row_spec(c(3,8,11),hline_after=TRUE) %>%
  row_spec(c(1,9,12), bold=T) %>%
  #kableExtra::kable_styling(latex_options = "scale_down") #%>% 
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8) # ensures the table isn't "floated" and pins it to the current location
```

\end{landscape}

### Data analysis

Differential expression analysis was performed using the 'edgeR' (v4.0.16) and 'limma' (v3.58.1) R packages. Lowly expressed genes (less than 10 counts and based on CPM as per recommended limma protocols) were filtered out if they were not expressed in at least four samples (chosen based on the sample size of the smallest ILD subtype, IPAF). The total number of genes in the final count matrix was 24,990. Genes were TMM-normalized and examined for outliers using multidimensional scaling (MDS) and PCA plots. Design matrices containing variables of interest were used to test for differential expression, with a term added into the formula to control for race as a confounding variable (Figure \@ref(fig:rnaseqPCA)). 

Survival analysis was performed using the elastic net implementation of Cox proportional hazards models in the 'glmnet' package (v4.1-8). Repeated 10-fold cross-validation was performed on all (100x repeat) or stratifications (50x repeat) of the centred and scaled TMM-normalized RNA-seq data. Using the C-index as a performance metric, the repeat with the C-index closest to the mean C-index of all repeats was selected as the final model. Validation datasets GSE28042 [@herazo-maya_peripheral_2013] and GSE93606 [@molyneaux_host-microbial_2017] were downloaded from GEO. As there were genes from the whole blood RNA-seq matrix missing from the validation datasets and vice versa, sparse matrices of each validation dataset were created for compatibility with the glmnet 'predict' function to account for missing variables. 

## Results

### The peripheral transcriptome is distinct between ILD subtypes and radiological patterns

We first examined differences in the peripheral transcriptome of specific ILD subtypes against each other. After controlling for race as a confounding variable (Figure \@ref(fig:rnaseqPCA)), IPF and SSc-ILD were the most transcriptionally distinct subtypes with 165 DEGs at an FDR<0.05, a majority of which were associated with interferon (IFN) signaling (Figure \@ref(fig:degild)A,C-D, Table \@ref(tab:ipfvsssc), Table \@ref(tab:ipfsscpathway)). No DEGs were identified between IPF and HP (Figure \@ref(fig:degild)B), but two DEGs at FDR<0.05 were identified between IPF and IPAF and 710 between SSc-ILD and IPAF, though these comparisons were limited by the small number of IPAF samples (n=4) (Table \@ref(tab:ILDdeg)). We further investigated DEGs between radiological subtypes, and found that fibrotic HP patterns were distinct from NSIP patterns with 16 DEGs as FDR<0.05 (Figure \@ref(fig:raddeg)); similar genes were upregulated when compared to UIP, though these were not significant (Table \@ref(tab:degrad)).

(ref:degild-cap) \textbf{SSc-ILD is transcriptionally distinct from other ILD subtypes.} Volcano plots of differentially expressed genes between (A) IPF and SSc-ILD and (B) IPF and HP. (C) Expression of interferon- and cytotoxic-associated genes in ILD subtypes. Statistical significance of blood RNA expression was corrected using false discovery rate (FDR), with significance indicated by FDR<0.01 (\*\*\*) and FDR<0.05 (\*\*). (D) Gene-concept plot of selected KEGG and Reactome pathways associated with up- and down-regulated SSc-ILD genes when compared to IPF. 

```{r degild, eval = T, out.width='1.0\\linewidth',echo=FALSE, fig.cap="(ref:degild-cap)", fig.scap = "SSc-ILD transcriptome", out.extra='', fig.align='center'}
knitr::include_graphics("./Figures/BloodRNAseq/Fig1_Differential_Expression_v1.pdf")
```

(ref:raddeg-cap) \textbf{Radiological patterns have distinct peripheral transcriptomes.} Volcano plots of differentially expressed genes between (A) fHP and NSIP and (B) fHP and UIP. (C) Expression of top DEGs in fHP. Satistical significance of blood RNA expression was corrected using false discovery rate (FDR), with significance indicated by FDR<0.01 (\*\*\*), FDR<0.05 (\*\*), and FDR<0.10(\*).

```{r raddeg, eval = T, out.width='1.0\\linewidth',echo=FALSE, fig.cap="(ref:raddeg-cap)", fig.scap = "Radiological transcriptome", out.extra='', fig.align='center'}
knitr::include_graphics("./Figures/BloodRNAseq/radiology_deg_v1.pdf")
```

### Genes are associated with PPF at baseline in SSc-ILD and NSIP patients

To investigate the association of the peripheral transcriptome at baseline with future development of PPF, we used the INBUILD criteria to classify whether or not patients had a PPF phenotype within 2 years. We did not identify any DEGs associated with PPF status in all ILD subtypes, although we identified a small number of genes significantly associated with INBUILD criteria such as relative decline in FVC of >5%, >10%, and radiological progression within 2 years (Figure \@ref(fig:ppfdeg), Table \@ref(tab:ppfdegtab)). We also did not identify any genes associated with PPF status when the analysis was stratified for patients with IPF, HP, UIP, or fHP, although we identified \textit{THBS3-AS1} expression associated with relative decline in FVC% predicted of >10% within 2 years in IPF patients (Figure \@ref(fig:ppfdegipf)). When the analysis was stratified within patients with SSc-ILD or NSIP, we identified increased \textit{LINC02490} expression as significantly associated with PPF status in both groups (FDR=0.003 for SSc-ILD, FDR=3.75\times 10\textsuperscript{-6}; Figure \@ref(fig:sscnsipppf)A-E, Table \@ref(tab:ppfdegtab)). Analysis of individual PPF criteria further identified \textit{LINC02470} and \textit{CLEC2A} as being significantly associated with radiological progression within 2 years in SSc-ILD but not NSIP patients, and similarly \textit{AOX1} and \textit{PAQR9} in NSIP patients (Figure \@ref(fig:sscnsipppf)F,G, Table \@ref(tab:ppfradtab)). 

(ref:sscnsipppf-cap) \textbf{PPF is associated with LINC02490 expression in patients with SSc-ILD and NSIP.} Volcano plots of differentially expressed genes between (A) PPF and non-PPF in SSc-ILD and (B) NSIP, and (C) radiological progression within 2 years in SSc-ILD and (D) NSIP. Boxplots showing expression of (E) \textit{LINC02490} in NSIP and SSc-ILD, and genes associated with radiological progression within 2 years in (F) SSc-ILD) and (G) NSIP. Satistical significance of blood RNA expression was corrected using false discovery rate (FDR), with significance indicated by FDR<0.01 (\*\*\*) and FDR<0.05 (\*\*).

```{r sscnsipppf, eval = T, out.width='0.9\\linewidth',echo=FALSE, fig.cap="(ref:sscnsipppf-cap)", fig.scap = "PPF transcriptome in SSc-ILD and NSIP", out.extra='', fig.align='center'}
knitr::include_graphics("./Figures/BloodRNAseq/ppf_ssc_deg_v1.pdf")
```

### Peripheral gene signatures are associated with transplant-free survival

We next used the elastic net implementation of Cox proportional hazards models to examine whether multivariate gene signatures were associated with transplant-free survival. By using repeated 10-fold cross validation, we generated predictive survival models using different stratifications of our dataset (all ILD, IPF, and ILD excluding SSc-ILD due to its interferon-associated DEGs). The worst performing model was the IPF stratified model (C-index = 0.45 [0.42-0.49]), though the all ILD (C-index = 0.67 [0.66-0.68]; Figure \@ref(fig:ildsurvival)A, Table \@ref(tab:ildsurvivalgenes)) and ILD excluding SSc-ILD (C-index = 0.76 [0.74-0.78]; Figure \@ref(fig:notsscsurvival)A, Table \@ref(tab:notsscsurvivalgenes)) models showed good performance. To validate these models, we used IPF samples from two external datasets with available transplant-free survival metadata: GSE28042 (PBMC) and GSE93606 (whole blood). Predicted hazard ratios from the ILD survival model were able to significantly separate high- and low-risk groups as defined by tertiles or optimized cutpoints (Figure \@ref(fig:ildsurvival)B,C). Similar results were obtained with the survival model trained on ILD samples excluding SSc-ILD (Figure \@ref(fig:notsscsurvival)). 

(ref:ildsurvival-cap) \textbf{ILD survival model derived from the peripheral transcriptome.} An ILD survival model derived using repeated 10-fold cross-validation was trained on our ILD cohort, with risk groups stratified by optimized cutpoints (A). The model was validated on two external IPF datasets, GSE28042 and GSE93606, which are shown with different risk group stratification based on splitting the hazard ratios into tertiles (B) or using optimized cutpoints (C). Log-rank \textit{P}-values shown in black are calculated between high- and low- risk groups, while the \textit{P}-value shown in grey is calculated between all three risk groups in (B). Selected genes in the model stratified by high and low expression are shown in (D). 

```{r ildsurvival, eval = T, out.width='1.0\\linewidth',echo=FALSE, fig.cap="(ref:ildsurvival-cap)", fig.scap = "ILD survival model", out.extra='', fig.align='center'}
knitr::include_graphics("./Figures/BloodRNAseq/PPF_survival_v2.pdf")
```

## Discussion

In this chapter, we provide a comprehensive examination of the peripheral transcriptome in ILD patients, and provide the first between-subtype comparison of blood gene expression between various ILD subtypes. We identified a distinct interferon-associated gene signature in patients with SSc-ILD, genes associated with ILD progression in SSc-ILD and NSIP patients, and a gene signature separating high- and low- risk patients with ILD with respect to transplant-free survival. 

Few molecular comparisons between ILD subtypes exist in lung samples, and, to our knowledge, none exist in blood samples. Furusawa \textit{et al.} identified similar up- and down-regulated genes in HP and IPF when compared to control lung samples, in addition to unique genes in HP related to immune processes [@furusawa_chronic_2020]. As discussed in the systematic review and three other studies [@yang_gene_2007; @cho_systems_2011; @horimasu_gene_2017], unique lung molecular signatures are difficult to identify for the idiopathic NSIP diagnosis when compared to other ILDs, which is likely due to its heterogenous population given the presence of the radiological pattern NSIP in many types of CTD-ILD. Although we did not identify any DEGs associated with IPF or HP, we found that SSc-ILD whole blood showed up-regulation of IFN-associated genes such as \textit{MX1}, \textit{IFIT1}, \textit{IFIT2}, \textit{IFIT3}, \textit{MX1}, \textit{OAS1}, \textit{OAS2}, \textit{OAS3}, \textit{ISG15}, \textit{IFI44}, and \textit{IFI44L} when compared to IPF, the majority of which are associated with type I responses though cross-over exists with type II responses [@singhania........]. Prior profiling of PBMCs from patients with SSc-ILD corroborate our findings, showing up-regulation of type I IFN genes when compared to healthy controls [@tan...]. Previous scRNA-seq work comparing cell populations from explanted lung tissue specimens of patients with IPF and SSc-ILD revealed diverse IFN signaling, with IPF showing increased type II IFN (IFN-\gamma) signaling and SSc-ILD displaying increased type I IFN (IFN-\alpha, IFN-\beta) signaling in macrophages, cytotoxic T cells, and NK cells [@valenzi_disparate_2021], which was also identified in bulk transcriptomics of ILD lung [@christmann_association_2014]. Taken together, this suggests that peripheral type I IFN signatures are unique in SSc-ILD and differentiate them from other ILDs. 

In our comparison of peripheral transcriptomes for radiological patterns, we identified a number of neutrophil-associated genes expressed in fHP when compared to NSIP and UIP, although the results in the latter comparison did not pass significance after correcting for FDR. While \textit{CEACAM8} is expressed on neutrophil membranes [@skubitz...], \textit{MMP8} and \textit{LCN2} are both associated with neutrophil migration and the latter forms a complex with \textit{MMP9} which was up-regulated in fHP as well when compared to NSIP at an FDR of 0.15 [@lin, @guardado...]. We also found up-regulation of \textit{LTF}, which is secreted by neutrophils to attenuate inflammation and promote tissue repair [@kruzel]. This is in line with the pathology of HP, as MMP-8- and MMP-9-positive neutrophils have been found to infiltrate lungs of patients with HP and were positively correlated with the extent of lung fibrosis [@pardo_increase_2000]. However, lung infiltration by neutrophils is not limited to HP, as similar results have been identified in patients with IPF and bleomycin-induced fibrosis in mice [@obayashi, @warheit-niemi]. Hence, further work must be done to validate a peripheral neutrophil gene signature of fHP as a potential diagnostic biomarker. 

We could not identify any differentially expressed genes between patients that did or did not develop PPF using the INBUILD criteria. However, stratified analysis within SSc-ILD revealed increased \textit{LINC02470} expression as significantly associated with PPF, which was also found in stratified analysis of NSIP samples though unsurprising as 31 of the 35 SSc-ILD patients had a radiological NSIP pattern. When investigating individual INBUILD criteria, expression of \textit{LINC02470} and \textit{CLEC2A} were associated with radiological progression of fibrosis within 2 years. \textit{LINC02470} has been shown to enhance EMT in bladder cancer by neutralizing miR-143-3p in order to rescue SMAD3 expression, which in turn activates TGF-\beta signaling [@huang...], while \textit{CLEC2A} is a skin-associated transcript expressed by NK and myeloid cells [@spreu...]. Furthermore, decline in FVC% predicted of >10% within 2 years in IPF patients was associated with decreased \textit{tHBS3-AS1}, which overlaps with miR-92b, a regulator of the TGF-\beta pathway; knockout of miR-92b aggravates kidney fibrosis [@Yang...]. 

Our work has certain limitations. While type I interferons have been explored as a biomarker in SSc-ILD [@kakkar...], the 



- ssc-ild vs other ild is not difficult diagnostically
- ssc-ild interferon relationship
- ssc-ild nk cell relationship

- fHP shows neutrophil signatures (LCN2, MMP8, CEACAM8); KREMEN1 blocks Wnt signaling

- ppf linc02470, commentary repeating what ryerson said about ppf in that different ppf criteria don't necessarily agree and suggest that the true relationship is radiological progression

- limitations regarding survival: 15/27 IPF patients censored, which is 15/25 of the total censored; need more peripheral transcriptional profiling done on other ILD subtypes 
- herazo-maya commentary: it doesn't work on our cohort, works well on the other two cohorts...

- COX15?




